---
output: github_document
bibliography: "REFERENCES.bib"
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# icuscoreR

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/icuscoreR)](https://CRAN.R-project.org/package=icuscoreR)
<!-- badges: end -->

icuscoreR is intended to collate `R` functions which apply clinical scoring systems to ICU patient data.

Currently a set of function calculate the Acute Physiology and Chronic Health Evaluation
(APACHE) III score (@Knaus1991-be) for patient data pertaining to the first 24 hours of
admission to an intensive care unit.

The `R` Data Frame must contain the field names described in the
Australia New Zealand Intensive Care Society (ANZICS) Adult Patient Data (APD)
Data Dictionary (@apddatadictionary). All field names are case insensitive, including a mandatory `IcuEpisodeId` field. `HI` and `LO` suffixes on field names such as `HRHI` (the highest heart rate in 24 hours from ICU admission) are optional, so `HRHI` can be `HR` if only 1 extreme is available, or if there are multiple rows of data for each `IcuEpisodeId`.

For greater detail, read `vignette(apache3)`.

Extensive and clearly written resources regarding variable collection, nomeclature and units of measure used by this package are available, hosted by ANZICS at
[ANZICS Data Collection Tools](https://www.anzics.org/data-collection-tools/).

## Installation

You can install the development version of [icuscoreR on GitHub](https://github.com/mcshaz/icuscoreR) with:

``` r
# install.packages("pak")
pak::pkg_install("mcshaz/icuscoreR")
```

## Example

See the `vignette(apache3)` for an in depth explanation of the following example

```{r example}
library(icuscorer)
data("ap3_appendix_example")
ap3_patient_data <- add_miss_ap3_cols(ap3_appendix_example)
ap3.scores <- apache3(ap3_appendix_example)
ap3.comorb.scores <- apache3_comorb(ap3_appendix_example)
data("ap3_appendix_gas")
ap3.abg.scores <- apache3_resp(ap3_appendix_gas)

ap3.scores <- merge(ap3.scores, ap3.abg.scores, by = "icuepisodeid")
ap3.scores <- merge(ap3.scores, ap3.comorb.scores, by = "icuepisodeid")
# Calculating the total Apache III score using
ap3.scores$totalap3 <- rowSums(ap3.scores[,endsWith(colnames(ap3.scores), 'ap3score')], na.rm = TRUE)
ap3.scores
```

*Note*

- At present the use cases for this package seem too niche to publish it to CRAN
, but please feel free to create an issue if you would prefer this to simplify
your workflow.
- Thank you to bgulbis who wrote and published [icuriskr](https://github.com/bgulbis/icuriskr).
At the time of writing, I believe `icuriskr` is a stale repository, including a
dependency on the `ICD` (International Classification of Diseases) `R` Package
which has been removed from CRAN, and also beware `icuriskr` contains a minor 
error in the calculation of pH component of the Apache III score
- If you believe other (adult or paediatric) scores should be added
(APACHE2, ANZROD, PIM2, PIM3, Trauma Severity Score, SOFA ...) please feel free to create an
issue (+/- pull request).

## Design Principles
- Consider performance and while maintaing readability, try to keep functions
vectorised (e.g. avoid `dplyr::rowwise`). It is conceivable these functions
could be executed on a multinational ICU database with 100&nbsp;000 rows of patient
data (and many fold more Arterial Blood Gas results).
- Consider different use cases and database structures
- Reference clear and publicly available data dictionaries
- Generate composite scores where feasible - the researcher using the score may
wish to include or exclude certain components because that is the exact component
under investigation
- Differentiate NA from a score of 0 even if 0 is the final score contributed

## Contributing
Contributions are welcome. However before submitting a pull request please:

- Create a Github issue first.
- Try and minimise dependencies beyond those existing (e.g. `DPLYR`) unless
there is a [reasonable case to do so](https://r-pkgs.org/dependencies-mindset-background.html).
- Please include documentation(`roxygen2`) and tests (`testthat`) with the pull
request.
- Code style follows the [Tidyverse style guide](https://style.tidyverse.org/)

## References
